Supplementary Information for RVD2: An ultra-sensitive variant detection model for low-depth heterogeneous next-generation sequencing data

### 2015-03-16 doc/bioinformatics2014/revision2/document.tex
Replace/add folder names in comments for each figure in main text and supplement with updated folder.
Submitted revision of bioinformatics RVD2 paper

\section{Results:}We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples. We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depth and minor allele frequency. We apply our model and identify fifteen mutated loci in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.

@@

\begin{equation}\label{eqn:germ_test}
	\Pr(\mu_j^{\text{control}}  \geq \tau ) \approx \frac{1}{N_{\text{MH}}} \sum_{k=1}^{N_{\text{MH}}} \mathbbm{1}_{\tilde{\mu}_{jk}^{\text{control}} \geq \tau } > 1-\alpha.
\end{equation}

@@

\begin{equation}\label{eqn:fisher_combined}
	X_j^2 = -2 \sum_{i=1}^N \ln(p_{ji}).
\end{equation}

@@

We measured the computational time for RVD2 varying the number of Gibbs sampling steps and the median read depth for the 400bp synthetic data data set. 
Briefly, on a 2.4GHz processor it took $\sim$13 min per 1000 Gibbs samples to fit the model. 
The computational time is independent of the median read depth due to the model structure; the same performance was observed for a median read depth of 130 and 40,000. 
As stated previously, due to the independence structure of the model, we are able to perform some of the sampling steps in parallel greatly decreasing the computational time.
The memory requirement is roughly the size of the gene sequence times the number of Gibbs samples.
Complete timing results are shown in Supplementary Section 8.

@@

We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data set by the false discovery rate and sensitivity/specificity. 
Among these algorithms, Samtools and GATK are optimized for homogeneous samples, while RVD, VarScan2-somatic, Strelka and muTect are have good performance to call variants in heterogeneous samples. 
In research applications, the false discovery rate is a more relevant performance metric because the aim is generally to identify interesting variants for followup. 
The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in correctly calling all of the positive variants and none of the negatives. 
GATK, Varscan2, Strelka and muTect are only able to make use of one case and one control sample, so we provide results of RVD2 with the same data set ($N=1$) for comparison.

@@

% from folder 'rvd2\results\2013-09-19_operating_characteristics'
% rvd2 data are from folder '2013-11-12_one_replicate_optT_with_chi','2013-11-12_one_replicate_T_zero_with_chi2','2013-11-15_six_replicates_optT_with_chi2','2013-11-15_six_replicates_T_zero_with_chi2'
% from folder 'rvd2\results\2013-09-19_operating_characteristics'

@@

\vspace{-10pt}
\subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca}
\begin{figure*}[!hbpt]
\centering
\vspace{-10pt}
\includegraphics[width=0.9\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
\caption{Estimated minor allele fraction for germline and somatic mutations called by RVD2 in the 59kbp PAXIP1 gene. Blue diamonds ($\diamond$) indicates germline mutations, where $\mu^{\text{control}}$ is significantly different from the reference sequence. Red stars (*) indicates somatic mutations, where $\mu^{\text{case}}$ is significantly different from $ \mu^{\text{control}}$. The  vertical lines represent 95\% credible interval around posterior mean MAF. Ten positions are common population SNPs according to dbSNPv138, and the identities are shown below the positions.}
\vspace{-10pt}
\label{fig:brca_MAF}
\end{figure*}
% figure 5 generated from 2014-03-24_germline_and_somatic_test_correction_no_subclassification

@@

% figure 6 generated from 2014-03-12_dot_plot_mutation_detection_result_HCC1187_dataset

@@

\vspace{-5pt}
\section*{Acknowledgements}
P.F. and Y.H. were supported by seed funding from Worcester Polytechnic Institute. Y.H. was supported by PhRMA Foundation Informatics Grant 2013080079.

>>>>

\section{Results:}We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples. We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depth and minor allele frequency. We apply our model and identify fifteen mutated loci in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are likely loss-of-heterozygosity events.

@@

\begin{equation}\label{eqn:germ_test}
	\Pr(\mu_j^{\text{control}}  \geq \tau ) \approx \frac{1}{N_{\text{MwG}}} \sum_{k=1}^{N_{\text{MwG}}} \mathbbm{1}_{\tilde{\mu}_{jk}^{\text{control}} \geq \tau } > 1-\alpha.
\end{equation}

@@

\begin{equation}\label{eqn:fisher_combined}
	\chi_j^2 = -2 \sum_{i=1}^N \ln(p_{ji}).
\end{equation}

@@

We measured the computational time for RVD2 varying the number of Gibbs sampling steps and the median read depth for the 400bp synthetic data data set. 
Briefly, on a 2.4GHz processor it took $\sim$13 min per 1000 Gibbs samples to fit the model. 
The computational time is independent of the median read depth due to the model structure; the same performance was observed for a median read depth of 130 and 40,000. 
As stated previously, due to the independence structure of the model, we are able to perform the sampling step for each location in parallel greatly decreasing the computational time.
The memory requirement is roughly the size of the gene sequence times the number of Gibbs samples.
Complete timing results without parallelization are shown in Supplementary Section 8.

@@

We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data set by the false discovery rate and sensitivity/specificity. 
Among these algorithms, Samtools and GATK are optimized for homogeneous samples, while RVD, VarScan2-somatic, Strelka and muTect have good performance to call variants in heterogeneous samples. 
In research applications, the false discovery rate is a more relevant performance metric because the aim is generally to identify interesting variants for followup. 
The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in correctly calling all of the positive variants and none of the negatives. 
GATK, Varscan2, Strelka and muTect are only able to make use of one case and one control sample, so we provide results of RVD2 with the same data set ($N=1$) for comparison.

@@

% from folder 'rvd2\results\2013-09-19_operating_characteristics'
% rvd2 data are from folder '2013-11-12_one_replicate_optT_with_chi','2013-11-12_one_replicate_T_zero_with_chi2','2013-11-15_six_replicates_optT_with_chi2','2013-11-15_six_replicates_T_zero_with_chi2'
% For revisions, Figure 3 and Figure 4 are generated from:
% from folder 'fzhang\Research\rvd\results\2015-03-09_run_RVD_replicates_thresh_SS_FDR_downsample_depth_100'
% from folder 'fzhang\Research\rvd\results\2015-03-03_run_Samtools_SS_FDR_downsample_depth_100'
% from folder 'fzhang\Research\rvd\results\2015-03-03_run_GATK_SS_FDR_downsample_depth_100'
% from folder 'fzhang\Research\rvd\results\2015-03-03_run_MuTect_SS_FDR_downsample_depth_100'
% from folder 'fzhang\Research\rvd\results\2015-03-03_run_Strelka_SS_FDR_downsample_depth_100'
% from folder 'fzhang\Research\rvd\results\2015-03-09_run_VarScan2_somatic_SS_FDR_downsample_depth_100'
% from folder 'fzhang\Research\rvd\results\2015-03-03_run_VarScan2_SS_FDR_downsample_depth_100'
% from folder 'fzhang\Research\rvd\results\2015-02-23_run_rvd2_SS_FDR_downsample_depth_100'

@@

\vspace{-10pt}
\subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca}
\begin{figure*}[!hbpt]
\centering
\vspace{-10pt}
\includegraphics[width=0.9\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
\caption{Estimated minor allele fraction for germline and somatic mutations called by RVD2 in the 59kbp PAXIP1 gene. Blue diamonds ($\diamond$) indicates germline mutations, where $\mu^{\text{control}}$ is significantly different from the reference sequence. Red stars (*) indicates somatic mutations, where $\mu^{\text{case}}$ is significantly different from $ \mu^{\text{control}}$. The  vertical lines represent 95\% credible interval around posterior mean MAF. Ten positions are common population SNPs according to dbSNPv138, and the identities are shown below the positions.}
\vspace{-10pt}
\label{fig:brca_MAF}
\end{figure*}
% figure 5 generated from '2014-03-24_germline_and_somatic_test_correction_no_subclassification'
% figure 5 generated from 'fzhang\Research\rvd\results\2015-02-25_MubarPlot_expanded_PAXIP1' for revisions.

@@

% figure 6 generated from '2014-03-12_dot_plot_mutation_detection_result_HCC1187_dataset'
% figure 6 generated from 'fzhang\Research\rvd\results\2015-02-25_dotPlot_expanded_PAXIP1' for revisions.

@@

\vspace{-5pt}
\section*{Acknowledgements}
P.F. was supported by seed funding from Worcester Polytechnic Institute. Y.H. and F.Z. were supported by PhRMA Foundation Informatics Grant 2013080079.

### 2015-03-16 doc/bioinformatics2014/revision2/supplementary_info.tex
Replace/add folder names in comments for each figure in main text and supplement with updated folder.
Submitted revision of bioinformatics RVD2 paper

\section{Histogram of $ \hat{\mu}_j $ for somatic mutations called by rvd2 in the 44kbp paxip1 gene from chr7:154738059 to chr7:154782774 in hcc1187 dataset}
\begin{figure}[H]
\begin{center}
\includegraphics[height=1\textheight]{pdf_figs/HCC1187_histogram_pos.pdf}
\caption{Histogram of $ \hat{\mu}_j $ for positions where $\mu^{\text{case}}$ is significantly higher than $ \mu^{\text{control}}$, namely Pr$ (\Delta \hat{\mu}_j> \tau) > 1- \alpha $, where $ \tau=0, \alpha=0.05 $.} 
\label{fig:hist_pos}
\end{center}
\end{figure}

@@

@@

\includepdf[pages={1-},scale=1]{pdf_figs/position_lookup_chart_all.pdf}

\section{Timing Results}

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.7\textwidth]{pdf_figs/timing.pdf}
\caption{Computational timing measurements for RVD2 on 400bp synthetic DNA data.}
\end{center}
\end{figure}

>>>>

\section{Histogram of $ \hat{\mu}_j $ for somatic mutations called by rvd2 in the 59kbp PAXIP1 gene from chr7:154735400 to chr7:154794682 in HCC1187 dataset}
\begin{figure}[H]
\begin{center}
\includegraphics[height=1\textheight]{pdf_figs/HCC1187_histogram_pos.pdf}
\caption{Histogram of $ \hat{\mu}_j $ for positions where $\mu^{\text{case}}$ is significantly higher than $ \mu^{\text{control}}$, namely Pr$ (\Delta \hat{\mu}_j> \tau) > 1- \alpha $, where $ \tau=0, \alpha=0.05 $.} 
\label{fig:hist_pos}
\end{center}
\end{figure}

@@

% Figure 5 and Figure 6 in supplementary are from folder '\fzhang\Research\rvd\results\2015-03-04_plot_posterior_mu_supp_PAXIP1'

@@

% Supplementary Table 1 is from folder '\fzhang\Research\rvd\results\2015-02-25_dotPlot_expanded_PAXIP1'
\includepdf[pages={1-},scale=1]{pdf_figs/position_lookup_chart_all.pdf}

\section{Timing Results}
% Figure 7 in supplementary is from folder '\fzhang\Research\rvd\results\2015-03-09_timing_rvd2_different_ngibbs'
\begin{figure}[H]
\begin{center}
\includegraphics[width=0.7\textwidth]{pdf_figs/timing.pdf}
\caption{Computational timing measurements for RVD2 on 400bp synthetic DNA data.}
\end{center}
\end{figure}
